---
title: "Project 2 EDA"
author: "Tejas Rawal"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

```{r include=FALSE}
# The package "ezids" (EZ Intro to Data Science) includes a lot of the helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
library(ezids)
 
# some of common options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
# knitr::opts_chunk$set(warning = F, results = "markup", message = F)
knitr::opts_chunk$set(warning = F, results = "markup", message = F)
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
# ‘scipen’: integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation.  Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than ‘scipen’ digits wider.
# use scipen=999 to prevent scientific notation at all times
```

```{r data init}
#loading packages 
library(ezids)
library(ggplot2)
library(ggrepel)
library(gridExtra)
library(tibble)

#loading data 
NYweath <- data.frame(read.csv("data/NYC_weather_1869_2022.csv"))

#converting to R date format and adding columns for day, month, and year
NYweath$DATE <- as.Date(NYweath$DATE)
NYweath$day <- format(NYweath$DATE, format="%d")
NYweath$month <- format(NYweath$DATE, format="%m")
NYweath$year <- format(NYweath$DATE, format="%Y")

#converting temperature observations to numerical
NYweath$TMAX <- as.numeric(NYweath$TMAX)
NYweath$TMIN <- as.numeric(NYweath$TMIN)
NYweath$TAVG <- as.numeric(NYweath$TAVG)
NYweath$year <- as.numeric(NYweath$year)

#Making month a factor
NYweath$month <- as.factor(NYweath$month)

# subset data to desired variables
NYweath_sub <- subset(NYweath, select = c(DATE, day, month, year, TMAX, TMIN, TAVG, PRCP, SNOW)) 

#creating a subset for 1900 on
NYweath_00 <- subset(NYweath_sub, year > 1899)
xkabledplyhead(NYweath_00)
```


1. Logistic regression to predict rainy day based on TMAX, TMIN, MONTH, DAY?  
2. Can we predict the month based on weather variables (KNN)?  
2. Correlation between Air Quality and Weather variables  

First, we format our data to create a variable representing whether it rained on a given day, starting :

```{r rainy days data}
# add column based on condition of rain on a given day
NYweath.rain <- NYweath_00
NYweath.rain$rained <- ifelse(NYweath.rain$PRCP > 0.0, 1, 0)
NYweath.rain$rained <- as.factor(NYweath.rain$rained)
xkablesummary(NYweath.rain)
xkabledplyhead(NYweath.rain)
```

First, let's create a two-way contingency table To study the effects on rainy day by the factor month and make sure there are no cells of zero frequencies.  

```{r crosstable}
rainy_vs_month = xtabs(~ rained + month, data = NYweath.rain)
rainy_vs_month
```

We can then quickly run a chi-squared test to see if the two are independent (or same frequency distribution).  
```{r chisq}
chisq.result = chisq.test(rainy_vs_month)
chisq.result
```

Based on the result, the factor variables `rained` and `month` are dependent - there is a statistically significant relationship between the two.  

### The Logistic Model

Let's jump to the logistic regression to predict a precipitation event on a given day:

```{r logit attempt}
rainedLogit <- glm(rained ~ TMAX + TMIN + month, data = NYweath.rain, family = "binomial")
summary(rainedLogit)
xkabledply(rainedLogit, title = paste("Logistic Regression :", format(formula(rainedLogit)) ))
```

All the coefficients are found significant (small p-values) except for the months of February, March, and April. TMAX has a negative effect on precipitation while TMIN has a positive effect. The months with significant coefficients all have a negative effect on the rain outcome for a given day.

Let's obtain the growth/decay factors for each explanatory variables. The factors are the exponentials of the coefficients:   

```{r growthDecayFactors, results='markup', collapse=F}
expcoeff = exp(coef(rainedLogit))
xkabledply(as.table(expcoeff), title = "Exponential of coefficients in rained Logit Reg")
```

### Logistic Model Evaluation
